Reasonably Programmable Literal Notation

### 2018-02-19 pldi18/rebuttal.txt

Thank you for the detailed reviews! We respond to the main questions and concerns below, then address additional questions beyond 700 words.

RA asked for more detailed examples in order to assess ease-of-use from the TLM provider’s perspective. The paper spoke to this perspective by pointing out that TLM parse functions are quite ordinary -- they take strings and produce ASTs represented as ordinary values -- so they can be generated using a parser generator of the provider’s choice, e.g. lex/yacc or a parser combinator library. Ease-of-use from the provider’s perspective is substantially a function of this choice. Including, e.g., a yacc grammar in the paper body would have consumed a large amount of space without conveying information relevant to the claims of the paper, which revolve around reasoning principles from the TLM client’s perspective.

The paper does provide a complete example -- Figure 4 shows the full definition of $html and two example applications. The only omission was a straightforward helper function that was instead detailed in the text. This example also spoke to this question of ease-of-use by demonstrating how providers can repurpose off-the-shelf parsers.

RB and RC expressed a more general desire for more examples. The prior work on TSLs presented a variety of examples and an empirical study that taxonomized additional possibilities, so in Sec1 we explicitly deferred to that paper in this regard. The case studies in the present paper were chosen to support our specific claims in consideration of the page limit. We will include additional examples in the final supplement.

We use “literal notation” to mean “notation for terms within a specified parametric type family” and will make this explicit in the final version.

RA said “I don't agree with the (implicit) argument in the paper that expansions
examined by a tool (e.g., IDE) are more abstract than expansions examined by a programmer”. The tool needs only examine the proto-expansion at the lowest level. It can then pass along only the type annotation and segmentation (a finite set) to other services, and ultimately to the programmer. The service/programmer can then reason about, e.g., the type of expression that is expected at a splice site. By analogy, a function’s type determines the type of its argument without requiring that the programmer inspect how that argument is used inside the function.

RA questioned the progression through simple TLMs before arriving at parametric TLMs. Simple TLMs are certainly not themselves practical (further discussion below the fold). Instead, they serve a pedagogical purpose: they allow us to tackle all five reasoning principles without bringing in the full complexity of the ML module system. Moreover, the impracticalities revealed by considering simple TLMs nicely motivate the introduction of parametric TLMs, which is why that section may seem concise relative to its practical importance.

RC says “all that TLMs do in this respect is to tag every “$” keyword with a nominal type”. The type need not be nominal (e.g. structural types like arrows are supported). Moreover, each spliced segment also has a type.

RC suggested that strict capture avoidance discipline we settled on prevents the expression of, e.g., regex backreferences. This is not the case -- TLMs can send values into spliced functions by application. Reason3 introduces concise function notation -- x => e -- that decreases the cost of this technique. The characteristic argument of this paper is that simpler reasoning principles justify these small syntactic costs.

RD wonders why we define the mechanism using arbitrary parse functions. Note that the reasoning principles we formulate would be equally relevant to designs that privilege a particular syntax definition formalism. Regarding AST representation, the ML type system is not powerful enough to express an intrinsically typed AST in the parametric setting, but if it were, the necessary invariants are exactly those maintained by proto-expansion validation.

RA and RB asked about inferring TLM parameters. It should be possible to infer those type parameters that appear in the TLM’s type annotation by unification at type-constrained  application sites. However, it would violate the Typing criterion to use the expansion itself to infer the type parameters. This might be acceptable if the editor reveals inferred parameters upon request, but we leave that to future work on editor design.

--- (additional responses below go beyond 700 words) --

RA brought up properties that clients cannot reason about abstractly, e.g. non-termination. However, we claim only to support the sort of abstract reasoning otherwise possible in the language, i.e. reasoning supported by the type system.

Several reviewers asked why we used “maps \hat{x} to \hat{x} ~> x” and similar instead of “maps \hat{x} to x”. This was an admittedly awkward way to show what the mapping connective looks like (“~>”). We will make this less awkward in the final version.

RA said “it is not clear why it is necessary to have $\hat{\Gamma} = \langle \mathcal{G}; \Gamma \rangle$ [...]” and RC had related questions. The point is that \hat{x} might be shadowed, which will cause it to be remapped, but we still need to track the “old” mapping (assumed unique by alpha-renaming in the XL) in \Gamma because it may have appeared elsewhere in the expansion. Figure 8 and the discussion around it gives such an example.

RA says that “one would need to change the variables in $\Delta$ and $\Gamma$ --- but they have already been picked”. Right -- an implementation strategy that would pre-empt the need to “backtrack” is one that generates fresh variables relative to \Delta/\Gamma_{app} each time \Delta/\Gamma is extended. The formalism is not meant to uniquely determine an implementation strategy, only to constrain it.

RC asks “is it not redundant then to have separate Typ and UTyp grammars”. The two grammars differ in that Typ contains variables and UTyp contains identifiers. The distinction is important because types can appear within expressions, including spliced expressions.

RC comments that “there is a reason we don’t write ASTs in our formalisms”. We used ABT notation deliberately because it is important to distinguish between terms that are and are not identified up to alpha variation. Moreover, UL terms are those that are exposed to the programmer, so we used conventional idioms, while XL terms are only ever manipulated as ABTs by parse functions, so we used ABT notation.

RC said that “markup is inconsistent between splicing using {[ ]} and just { }, within an HTML literal”. Those are different splicing forms -- the first for string splicing and the second for HTML splicing (see Figure 1 and accompanying text).

RD says “you couldn’t define useful pattern expansions using guards”. In OCaml, guards are not part of the syntax of patterns (they can appear only at the outermost level of a match rule). If guards were part of patterns, then the mechanisms already considered for expression TLMs would become relevant to the guard portion of a pattern expansion as well.

RC asked about the “fold” operation and RD asks “what is the `fold` pattern?” These are the introductory form for recursive types. Similarly, the “inj[\ell]” operation is the introductory form for sum types. We based our terminology on Harper’s PFPL, but the same or similar notation is common in other textbooks (e.g. Pierce) and in the literature.

RA says “I think that the paper drops some non-trivial information in the simplification and abbreviation of the proto_typ, proto_expr, and proto_pat types.” Yes, these datatypes were simplified as acknowledged in the text. In practice, we use the existing OCaml AST, with a special variable “__tlm_spliced” that takes a pair of number literals to represent spliced segments. We will mention this implementation detail in the final version.

The $grammar TLM in the supplement, and related definitions, were shown only to outline how a parser generator could be formulated as a mode of use of TLMs. A full parser generator implementation would span thousands of lines of code. There is certainly no requirement that the programmer use this method, and we expect in practice that existing parser generators will be used to generate parse functions.

### 2018-02-20 icfp18/syntax-icfp18.tex
initial port to ICFP18 format and some edits to the intro that preview my plan for the submission

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Cyrus packages
\usepackage{microtype}
\usepackage{todonotes}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\definecolor{light-gray}{gray}{0.95}

@@

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray},
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=~,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,syntax,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=-10pt, 
aboveskip=4pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\lstset{morecomment=[n]{/*}{*/}}

@@

@@

General-purpose programming languages typically define literal notation for only a small number of data structures, e.g. lists. This is unsatisfying because there are many other data structures for which literal notation might be useful, e.g. finite maps, regexes, HTML data, SQL queries, syntax trees and encodings of chemical structures. %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary form at any specified type or parameterized family of types. 

@@

Compared to existing approaches, including direct grammar extension systems and prior  macro-based parsing systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e. without examining the underlying expansion, about types and binding. The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal, which gives the locations of spliced subterms. 

@@

This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules. 
%This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles.

@@

% Put succinctly, we design a programming language in the ML tradition with a \emph{reasonably} programmable syntax.
\vspace{-10px}
\end{abstract}

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
\vspace{-3px}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few data structures. For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows two HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This language design practice is decidedly \emph{ad hoc}, in that it is easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. For example, 1) clients of a ``collections'' library might also want matrix, set and map literals following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar. Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat strings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure*}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\vspace{-10px}
\label{fig:urweb}
\end{figure*}

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms. Instead, there has been considerable interest for many years in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and SugarJ \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec. \ref{sec:existing-approaches} can all be used to define new literal notation.

\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e. by inspecting only the surface syntax, about such fundamental issues as types and variable binding when the program text is littered with  unfamiliar user-defined literal forms. Instead, programmers can only reason transparently, i.e. by inspecting the underlying expansion or the code responsible for producing the expansion. In some existing systems, even determining which extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''

\begin{figure}[t!]
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
\end{lstlisting}
\vspace{-10px}
\caption{Syntax extension systems produce unreasonable program text.}
\vspace{-7px}
\label{fig:K-dialect}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason-what}\footnote{This paper uses Reason 2.0 as 3.0 was released just before submission.} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework}. 
The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:

\begin{enumerate}[nolistsep,leftmargin=10pt]
\item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3? Might activating a new extension generate a conflicting expansion for the same form (i.e. is there syntactic ambiguity)?
\item \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the expansion of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

An approach that has made some headway against these problems is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. With this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against. Generalized literal forms syntactically subsume other literal forms because their bodies---the characters between the outer delimiters, here \li{[|} and \li{|]}---are left  unparsed by the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) With TSLs, Line 3 of our example would require a type annotation:
\begin{lstlisting}[numbers=none]
    let y : kquery = [|(!R)@&{&/x!/:2_!x}'!R}|]
\end{lstlisting}
The parse function associated with the \li{kquery} type would be invoked to parse and expand the literal body.

TSLs support reasoning abstractly about \textbf{Responsibility} and \textbf{Typing}, due to the type-driven dispatch mechanism, but the remaining reasoning principles were not fully and formally established in the prior work. Moreover, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml. These included non-trivial problems related to support for structural types (e.g. arrow types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.  Another problem was that parse functions and expansions were completely restricted from accessing libraries. Taken together, we found that TSLs as described were not suitable for integration into Reason/OCaml.

\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into Reason/OCaml and other full-scale statically typed languages. We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper and involve all of the aforementioned language features. The empirical study in the prior work on TSLs \cite{TSLs} supports the claim that a macro-based mechanism for defining new literal notation is broadly useful, so this paper is targeted at language designers that seek a deeper understanding of the reasoning principles at play, particularly as they relate to features common to ML and other full-scale statically typed languages (e.g. Haskell, Scala, Rust, etc.)

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% \documentclass[sigplan,review,10pt,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% \documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Cyrus packages
\usepackage{microtype}
\usepackage{todonotes}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}
\usepackage{wasysym}
\definecolor{light-gray}{gray}{0.95}

@@

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray},
stringstyle=\ttfamily\color{red},
mathescape=false,escapeinside={(~}{~)},
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=-6pt, morekeywords=[1]{try,spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,string,datatype,new,toast,notation,module,switch,where,using,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double,structure,of},classoffset=0, xleftmargin=0pt, 
aboveskip=4pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\lstset{morecomment=[n]{/*}{*/}}

@@

\setlength{\abovecaptionskip}{4pt plus 3pt minus 2pt} % Chosen fairly arbitrarily
\setlength{\belowcaptionskip}{-2pt plus 3pt minus 2pt} % Chosen fairly arbitrarily

@@

General-purpose programming languages typically define literal notation for only a small number of common data structures, e.g. lists. This is unsatisfying because there are many other data structures for which literal notation might be useful, e.g. hash tables, regular expressions, HTML data, SQL queries, syntax trees and chemical structures. There may also be different implementations of each of these data structures, e.g. with different performance characteristics, that could all benefit from common literal notation. %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper introduces \emph{typed literal macros (TLMs)}, which allow library providers to define new literal notation of nearly arbitrary design at any specified type or parameterized family of types. 

@@

Compared to existing approaches, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic ambiguity, and TLM clients can reason abstractly, i.e. without examining the underlying expansion, about types and binding. The system only needs to convey to clients, via secondary notation, the inferred \emph{segmentation} of each literal, which gives the locations and types of spliced subterms. 

@@

This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, pattern matching and ML-style modules. 
This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML. 

@@

% Put succinctly, we design a programming language in the ML tradition with a \emph{reasonably} programmable syntax.
\end{abstract}

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a select few data structures. For example, Standard ML supports list literals like \li{[x1, x2, x3]} in both expressions and patterns \cite{harper1997programming,mthm97-for-dart}, and many other general-purpose languages support substantially similar list literal notation. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Another example is Ur/Web, which extends the surface syntax of Ur (an ML-like language \cite{conf/pldi/Chlipala10}) with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15}. The example in Fig. \ref{fig:urweb} below shows two HTML literals, one that ``splices in'' a string expression delimited by \li{\{[} and \li{]\}} (Line 1) and the other an HTML expression delimited by \li{\{} and \li{\}} (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

\begin{figure}[h]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-4px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web \cite{conf/popl/Chlipala15}.}
\label{fig:urweb}
\vspace{-8px}
\end{figure}

This design practice, where the language designer privileges certain user-defined types with built-in literal notation, is decidedly \emph{ad hoc} in that it is easy to come up with other examples of data structures for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. For example, 1) clients of a ``collections'' library might want not just list literals, but also literal notation for matrices, finite sets, maps, hash tables and so on, following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler author might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of string injection attacks, which are amongst the most common and serious security vulnerabilities in web applications \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them could be parsed by some readily apparent type-specific grammar. Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than strings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms. Instead, there has been considerable and persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Omar et al's \emph{type-specific languages} (TSLs) \cite{TSLs}, and others that we will discuss in Sec.~\ref{sec:existing-approaches} can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix operators and type declaration forms, which are beyond the scope of this paper)\todo{normalize parens-period thing}.

\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e. by inspecting only the program as presented to the programmer, about such fundamental issues as types and variable binding when the program text is littered with  user-defined literal forms. Instead, programmers and editor services can only reason transparently, i.e. by inspecting the underlying expansion or the extension responsible for producing that expansion. In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
\newcommand*\Suppressnumber{%
  \lst@AddToHook{OnNewLine}{%
    \let\thelstnumber\relax%
     \advance\c@lstnumber-\@ne\relax%
    }%
}

\newcommand*\Reactivatenumber{%
  \lst@AddToHook{OnNewLine}{%
   \let\thelstnumber\origthelstnumber%
   \advance\c@lstnumber\@ne\relax}%
}

\begin{figure}[t!]
\begin{subfigure}[t]{0.48\textwidth}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={context,free,syntax}]
  context-free syntax 
    (* ...direct mods to the CFG... *)
  (* ... *)
  let w = compute_w();
  let x = compute_x(w);
  let y = `(!R)@&{&/x!/:2_!x}'!R`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the five reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t by KQueryParser 
    using module Helpers as Helpers;
  (* ... *)
  let w = compute_w();
  let x = compute_x(w);
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the grammar fixed and making the segmentation and the type and binding structure explicit.}
\label{fig:K-tsm-example}
\end{subfigure}
\caption{Literal notation for encodings of K queries}
\vspace{-10px}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{convert to Reason 3}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system like Camlp4 \cite{ocaml-manual} or Sugar*~\cite{erdweg2011sugarj,erdweg2013framework} (Line 1 suggestively uses the syntax for introducing a grammar extension from SugarHaskell \cite{erdweg2012layout}). 
The literal form on Line 6 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot easily answer questions like the following:

\begin{enumerate}[leftmargin=12px]
\item[\Circle] \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 6? Might activating a new extension generate a conflicting expansion for the same form (i.e. might there be a syntactic ambiguity)?
\item[\Circle] \textbf{(Segmentation)} Are the characters \li{x}, \li{R} and \li{2} on Line 6 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item[\Circle] \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 5, or might it capture an unseen binding of the same identifier in the expansion of Line 6?
\item[\Circle] \textbf{(Context Dependence)} If \li{w}, on Lines 4-5, is renamed, could that possibly break Line 6 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item[\Circle] \textbf{(Typing)} What type does \li{y} have? What type is each spliced expression expected to have?
\end{enumerate}

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}. Ideally, the programmer would be able to abstract in some analagous manner over the implementation of a notation. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.
